Insights Into Computational Reproducibility in Plant Pathology and a Way Forward

### 2020-12-07 analysis/paper/paper.Rmd
grammar and spelling corrections

---
title: "Insights Into Computational Reproducibility in Plant Pathology and a Way Forward"
author:
  - Adam H. Sparks:
      email: Adam.Sparks@usq.edu.au
      institute: USQ
      correspondence: true
  - Emerson Del Ponte:
      institute: UFV
  - Kaique dos Santos Alves:
      institute: UFV
  - Zachary Foster: 
      institute: OSU
  - Niklaus J Grünwald: 
      institute: ARS
institute:
  - USQ: University of Southern Queensland, Centre for Crop Health, Toowoomba, Queensland, Australia
  - UFV: Departmento de Fitopatologia, Universidade Federal de Viçosa, Brazil
  - OSU: Department of Botany and Plant Pathology, Oregon State University, Corvallis, OR, USA
  - ARS: Horticultural Crops Research Unit, USDA Agricultural Research Service, Corvallis, OR, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  officedown::rdocx_document:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
        - '--lua-filter=pagebreak.lua'
      tables:
        layout: autofit
        width: 1.0
        caption:
          style: Table Caption
          pre: 'Table '
          sep: ': '
        conditional:
          first_row: true
          first_column: false
          last_row: false
          last_column: false
          no_hband: false
          no_vband: true
bibliography: references.bib
csl: "../templates/TPP.csl" # Insert path for the bib-style
abstract: |
  Open research practices have been highlighted extensively during the last ten years in many fields of study as essential standards needed to promote transparency and reproducibility of scientific results. In fact, scientific claims can only be evaluated based on how protocols, materials, equipment and methods were described; data were collected and prepared; and analyses conducted. Openly sharing protocols, data and computational code are central for current scholarly dissemination and communication, but in many fields, including plant pathology, adoption of these practices has been slow. We randomly selected 300 articles published from 2012 to 2018 across 21 journals representative of the pathology discipline, and assigned them scores reflecting their openness and reproducibility. We found that most of the articles were not very open, failing to share data or code in a meaningful way. We propose that using open source tools for producing open and reproducible work and analysis is advantageous, benefiting not just readers, but the authors as well and provide ideas and tools to promote open, reproducible research practices among plant pathologists.
keywords: |
  open science; open data; reproducible research; FAIR data
highlights: |
  Open research practices are highlighted extensively in the broader scientific community. Our survey of 300 randomly selected articles from 2012 to 2018 indicates that we have work to do in this area as plant pathologists with most articles scoring low in openness and reproducibility.
---

@@

Modern plant pathological research has many facets given the array of disciplines and sub-disciplines currently involved.
Collectively, they contribute to increase our basic and applied knowledge of several aspects in pathogen biology and disease development to ultimately improve plant disease management.
Scientific research in the field varies from the purely observational or descriptive nature to inferential, based on experimental or simulation-derived data sets.
Whatever the case, research findings are verifiable based on how much of the research materials, processes and outcomes are made available beyond what is reported in the scientific article and the ability of others to make use of the methods and results.
These research findings include biological materials (isolates or strains), nucleic/protein sequences, experimental and simulated raw data annotations, drawings and photographs and statistical analysis code among other materials and data generated as a course of conducting research.

@@

In order for us to easily discuss the topic, we first must define what we mean so that we may clearly communicate.
Many of the terms used in this area have varying definitions that may or may not agree with each other. For instance, reproducible research is recently highlighted by many authors [@Baker2016a; @Iqbal2016; @Nature2016; @Patil2016; @Weissgerber2016; @Brunsdon2015; @Sweedler2015; @Fitzjohn2014; @Ioannidis2014; @Fidler2013; @Stodden2013] as an important issue.
The International Union of Pure and Applied Chemistry provides
definitions of repeatability in the "Compendium of Chemical Terminology -- Gold Book" as:

@@

Therefore our definition of reproducibility will be that the computer code and data are made freely available to others for verification and conducting alternate analyses or for use in instructional purposes. And that the software used are also easily obtainable and preferably open source to avoid licensing or other issues related to accessibility for end-users related to costs or non-standard file formats, etc.

In an effort to understand where as a discipline we stand as plant pathologists in regards to open science and reproducible research, we conducted a survey of a broad selection of articles to represent a broad swathe of publications to evaluate our status.

@@

We randomly selected 300 articles published from 2012 to 2018 across 21 journals (Table \@ref(tab:journals)) and assigned them scores reflecting their openness and reproducibility.
The journals represented both fundamental and applied areas as well as pathogen or other group specific publications.
A list of randomly selected pages was assigned to a randomized list of the 21 journals where the page number fell within an article for the given journal.
In cases where an article was not suitable, _e.g._, a review or otherwise not related to plant pathology, the next article was selected until a suitable article was found.
Notes regarding the selection of articles can be found in the file, XXXX, available in this paper's repository.
The pages list was numbered from page one and went to page 150.
This was done since some journals restart their numbering with each issue and also ensures that the journal is more likely to have a page number corresponding to the randomly generated value.
This also assumes that there is no effect or bias on reproducibility based on the time of year that an article was published, since most journals start with page number one at the beginning of the year.
The list of journals was saved from Google's sheets as a comma separated value (CSV) file for later use and distribution in a research compendium.

@@

Each journals was classified as to  whether they were completely open (TRUE), behind a pay wall (FALSE) or a combination (BOTH).
The five-year impact factor for 2018 for each journal was retrieved from InCites Journal Citation Reports, Clarivate Analytics and entered in a separate sheet in the Google sheets file.
This was downloaded, saved as a CSV file and is left-joined with the article notes using a `left_join()` from the 'dplyr' package [@Wickham2020] for analysis when the `import_notes()` command is executed.

Articles were classified as fundamental or applied and a note was made whether the article's primary focus was using or developing molecular techniques (TRUE/FALSE).
This classification was based on the evaluator's own judgement of the article.

@@

    * 1 - Available upon request to author

    * 2 - Online, but inconvenient or non-permanent (_e.g._, login needed, pay wall, FTP server, personal lab website that may disappear, or may have already disappeared)

@@

A total reproducibility score was calculated as the percent of total possible for a given paper,
_e.g._ some papers did not use specialized software therefore they were not scored for this category.

We recognise that the scoring methods were developed by the authors specifically for this exercise and are not entirely subjective.
We have based our scoring decisions on our own experiences in the field of plant pathology.
While many software packages may be widely used or "freely" available if one has an institutional licence, these types of software often hinder partners and interested parties that do not have these luxuries.
Thus, emphasis was given to using open source alternatives where possible over paid software packages.
The view of the authors is one that this is something to be encouraged as it allows the broadest possible audience to benefit from the work at the least cost.

@@

Bayesian mixed effect models were fit using the contributed package 'brms' [@Burkner2018] in the R environment  [@RCT2020] to evaluate the fixed effects of the publication (journal title), article class, molecular focus, five year journal impact factor, journal open access policy on the reproducibility index and year to test for changes over time in the reproducibility index.

@@

The random effects were year and assignee for all models excepting the model which tested for year effects.
In this model the random effects were the publication title and assignee.

@@

Forty-eight articles scored 0% out of a possible 100% for reproducibility.
The mean value was 31.2% and the median value was 33.3%.
One article scored 100%, or was classed as fully reproducible.

@@

Most articles did not make any computational methods available in any fashion with two, 0.007%, classing as "3", which was the highest score available.
Two hundred eighty seven, 96%, were classed as "0" and 9 articles, 0.03%, appeared to not use any computational methods.

@@

The most frequently cited software was SAS.
The remainder were mostly specialised software with the exception of Microsoft Excel, tied for 9th (Table \@ref(tab:su)).

@@

* Sparks et al. [-@Sparks2011; -@Sparks2014] provide models, data and code, (http://adamhsparks.github.io/Global-Late-Blight-MetaModelling/)
necessary to replicate model development and the subsequent the study on the effects of climate change on potato late blight.

* Lehner et al. [-@Lehner2017] provides data and a reproducible report that explain in details all steps of the analysis and the R codes for conducting a meta-analysis for assessing heterogeneity in relationship between white mold incidence and soybean yield and between incidence and soybean tied (https://emdelponte.github.io/paper-white-mold-meta-analysis/).

@@

A general work flow for producing academic research involves clearly defining a research question, obtaining data for testing the hypothesis, summarizing/analyzing and presenting data and results, and writing the manuscript.
Here we defined three levels of reproducibility which are also related with the evolution of computational methods and reproducible practices (Figure \@ref(fig:workflow-dia)).

A first level of reproducibility involves openly available research materials such as isolates or strains and/or nucleic acid sequences in public collections and citations for methods used.
A second level involves providing raw data and code as binary files (PDF or other non-text file) in supplemental materials, which do not allow prompt access to the data and running the code because of use of expensive commercial software or a pay-wall.
The highest level includes efforts to annotate structured raw data (FAIR -- Findable, Accessible, Interoperable and Re-usable) [@Wilkinson2016] and fully document the analysis using open source code which are deposited in public repositories and can be run by anyone following download of data and code.
The first level, as reported, is an essential step that is not substituted by the other practices and eventually researchers fail to provide sufficient description or correct citations.
In the next section we present standards and tools that can be used to ensure reproducibility.

When making your science more open and reproducible, methods, software used, this includes items such as R or Python packages that were directly used in the analysis or production of the paper, etc. should be cited properly.
Deposit and annotate biological materials with an herbarium or other repository.
Provide a full description of all equipment used, _e.g._ a Spectrum Technologies Watchdog 2700 weather station was used to record wind speed, direction, rainfall, temperature and relative humidity at one hour intervals.
This allows end users to identify what was used and identify the methods used more accurately.
Just as importantly, this acknowledges the contributions of others whose works were instrumental in your research.
This also helps ensure that you as a researcher are able to reconstruct what you have done since you will have good notes and documentation and be able to identify if something changes, _e.g._ a package version, what effect it had on your research.

The use of programming or scripting languages such as R, SAS or Python enable you as a scientist to keep very detailed records of what was computationally performed.
This is as opposed to using software such as spreadsheet programs like Excel, Google Sheets, Numbers, Calc or others which can be used for simple statistical analyses and visualization (but should not be) or other point-and-click software packages that do not enable you to keep an accurate record of the steps taken to import, format, visualise and analyse data.

@@

If steps are followed to make the data FAIR then it will be readable by humans and machines alike this will help support new discoveries and support further research.
In turn sharing data will lead to new citations for your work as other discover and use it.

>>>>

---
title: "Insights Into Computational Reproducibility in Plant Pathology and a Way Forward"
author:
  - Adam H. Sparks:
      email: Adam.Sparks@usq.edu.au
      institute: USQ
      correspondence: true
  - Emerson Del Ponte:
      institute: UFV
  - Kaique dos Santos Alves:
      institute: UFV
  - Zachary Foster: 
      institute: OSU
  - Niklaus J Grünwald: 
      institute: ARS
institute:
  - USQ: University of Southern Queensland, Centre for Crop Health, Toowoomba, Queensland, Australia
  - UFV: Departmento de Fitopatologia, Universidade Federal de Viçosa, Brazil
  - OSU: Department of Botany and Plant Pathology, Oregon State University, Corvallis, OR, USA
  - ARS: Horticultural Crops Research Unit, USDA Agricultural Research Service, Corvallis, OR, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  officedown::rdocx_document:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
        - '--lua-filter=pagebreak.lua'
      tables:
        layout: autofit
        width: 1.0
        caption:
          style: Table Caption
          pre: 'Table '
          sep: ': '
        conditional:
          first_row: true
          first_column: false
          last_row: false
          last_column: false
          no_hband: false
          no_vband: true
bibliography: references.bib
csl: "../templates/TPP.csl" # Insert path for the bib-style
abstract: |
  Open research practices have been highlighted extensively during the last ten years in many fields of study as essential standards needed to promote transparency and reproducibility of scientific results. Scientific claims can only be evaluated based on how protocols, materials, equipment and methods were described; data were collected and prepared, and; analyses conducted. Openly sharing protocols, data and computational code is central for current scholarly dissemination and communication, but in many fields, including plant pathology, adoption of these practices has been slow. We randomly selected 300 articles published from 2012 to 2018 across 21 journals representative of the pathology discipline and assigned them scores reflecting their openness and reproducibility. We found that most of the articles were not very open, failing to share data or code in a meaningful way. We propose that using open source tools for producing open and reproducible work and analysis is advantageous, benefiting not just readers, but the authors as well and provide ideas and tools to promote open, reproducible research practices among plant pathologists.
keywords: |
  open science; open data; reproducible research; FAIR data
highlights: |
  Open research practices are highlighted extensively in the broader scientific community. Our survey of 300 randomly selected articles from 2012 to 2018 indicates that we have work to do in this area as plant pathologists with most articles scoring low in openness and reproducibility.
---

@@

Modern plant pathological research has many facets given the array of disciplines and sub-disciplines currently involved.
Collectively, they contribute to increasing our basic and applied knowledge of several aspects of pathogen biology and disease development to ultimately improve plant disease management.
Scientific research in the field varies from the purely observational or descriptive nature to inferential, based on experimental or simulation-derived data sets.
Whatever the case, research findings are verifiable based on how much of the research materials, processes and outcomes are made available beyond what is reported in the scientific article and the ability of others to make use of the methods and results.
These research findings include biological materials (isolates or strains), nucleic/protein sequences, experimental and simulated raw data annotations, drawings and photographs and statistical analysis code among other materials and data generated as a course of conducting research.

@@

For us to easily discuss the topic, we first must define what we mean so that we may clearly communicate.
Many of the terms used in this area have varying definitions that may or may not agree with each other. For instance, reproducible research is recently highlighted by many authors [@Baker2016a; @Iqbal2016; @Nature2016; @Patil2016; @Weissgerber2016; @Brunsdon2015; @Sweedler2015; @Fitzjohn2014; @Ioannidis2014; @Fidler2013; @Stodden2013] as an important issue.
The International Union of Pure and Applied Chemistry provides
definitions of repeatability in the "Compendium of Chemical Terminology -- Gold Book" as:

@@

Therefore our definition of reproducibility will be that the computer code and data are made freely available to others for verification and conducting alternate analyses or for use in instructional purposes. And that the software used are also easily obtained and preferably open source to avoid licensing or other issues related to accessibility for end-users related to costs or non-standard file formats, etc.

In an effort to understand where as a discipline we stand as plant pathologists in regards to open science and reproducible research, we surveyed a broad selection of articles to represent a broad swathe of publications to evaluate our status.

@@

We randomly selected 300 articles published from 2012 to 2018 across 21 journals (Table \@ref(tab:journals)) and assigned them scores reflecting their openness and reproducibility.
The journals represented both fundamental and applied areas as well as pathogen or other group-specific publications.
A list of randomly selected pages was assigned to a randomized list of the 21 journals where the page number fell within an article for the given journal.
In cases where an article was not suitable, _e.g._, a review or otherwise not related to plant pathology, the next article was selected until a suitable article was found.
Notes regarding the selection of articles can be found in the file, XXXX, available in this paper's repository.
The pages list was numbered from page one and went to page 150.
This was done since some journals restart their numbering with each issue and also ensures that the journal is more likely to have a page number corresponding to the randomly generated value.
This also assumes that there is no effect or bias on reproducibility based on the time of year that an article was published since most journals start with page number one at the beginning of the year.
The list of journals was saved from Google's sheets as a comma-separated-value (CSV) file for later use and distribution in a research compendium.

@@

Each journal was classified as to  whether they were completely open (TRUE), behind a paywall (FALSE) or a combination (BOTH).
The five-year impact factor for 2018 for each journal was retrieved from InCites Journal Citation Reports, Clarivate Analytics and entered in a separate sheet in the Google sheets file.
This was downloaded, saved as a CSV file and is left-joined with the article notes using a `left_join()` from the 'dplyr' package [@Wickham2020] for analysis when the `import_notes()` command is executed.

Articles were classified as fundamental or applied and a note was made whether the article's primary focus was using or developing molecular techniques (TRUE/FALSE).
This classification was based on the evaluator's judgement of the article.

@@

    * 1 - Available upon request to the author

    * 2 - Online, but inconvenient or non-permanent (_e.g._, login needed, paywall, FTP server, personal lab website that may disappear, or may have already disappeared)

@@

A total reproducibility score was calculated as the per cent of total possible for a given paper,
_e.g._ some papers did not use specialized software therefore they were not scored for this category.

We recognise that the scoring methods were developed by the authors specifically for this exercise and are not entirely subjective.
We have based our scoring decisions on our own experiences in the field of plant pathology.
While many software packages may be widely used or "freely" available if one has an institutional licence, these types of software often hinder partners and interested parties that do not have these luxuries.
Thus, the emphasis was given to using open source alternatives where possible over commercially available, non-free software packages.
The view of the authors is one that this is something to be encouraged as it allows the broadest possible audience to benefit from the work at the least cost.

@@

Bayesian mixed-effect models were fit using the contributed package 'brms' [@Burkner2018] in the R environment  [@RCT2020] to evaluate the fixed effects of the publication (journal title), article class, molecular focus, five-year journal impact factor, journal open access policy on the reproducibility index and year to test for changes over time in the reproducibility index.

@@

The random effects were year and assignee for all models excepting the model which tested for year effects.
In this model, the random effects were the publication title and assignee.

@@

Forty-eight articles scored 0% out of a possible 100% for reproducibility.
The mean value was 31.2% and the median value was 33.3%.
One article scored 100% or was classed as fully reproducible.

@@

Most articles did not make any computational methods available in any fashion with two, 0.007%, classing as "3", which was the highest score available.
Two hundred eighty-seven articles, 96%, were classed as "0" and nine articles, 0.03%, appeared to not use any computational methods.

@@

The most frequently cited software was SAS.
The remainder were mostly specialised software except for Microsoft Excel, tied for 9th (Table \@ref(tab:su)).

@@

* Sparks et al. [-@Sparks2011; -@Sparks2014] provide models, data and code, (http://adamhsparks.github.io/Global-Late-Blight-MetaModelling/)
necessary to replicate model development and the subsequent study on the effects of climate change on potato late blight.

* Lehner et al. [-@Lehner2017] provide data and a reproducible report that explains in detail all steps of the analysis and the R codes for conducting a meta-analysis for assessing heterogeneity in the relationship between white mold incidence and soybean yield and between incidence and soybean tied (https://emdelponte.github.io/paper-white-mold-meta-analysis/).

@@

A general work flow for producing academic research involves clearly defining a research question, obtaining data for testing the hypothesis, summarizing/analyzing and presenting data and results, and writing the manuscript.
Here we defined three levels of reproducibility which are also related to the evolution of computational methods and reproducible practices (Figure \@ref(fig:workflow-dia)).

The first level of reproducibility involves openly available research materials such as isolates or strains and/or nucleic acid sequences in public collections and citations for methods used.
A second level involves providing raw data and code as binary files (PDF or other non-text files) in supplemental materials, which do not allow prompt access to the data and running the code because of use of expensive commercial software or a pay-wall.
The highest level includes efforts to annotate structured raw data (FAIR -- Findable, Accessible, Interoperable and Re-usable) [@Wilkinson2016] and fully document the analysis using open source code which are deposited in public repositories and can be run by anyone following the download of data and code.
The first level, as reported, is an essential step that is not substituted by the other practices and eventually researchers fail to provide sufficient description or correct citations.
In the next section we present standards and tools that can be used to ensure reproducibility.

When making your science more open and reproducible, methods, software used, this includes items such as R or Python packages that were directly used in the analysis or production of the paper, etc. should be cited properly.
Deposit and annotate biological materials with an herbarium or other repository.
Provide a full description of all equipment used, _e.g._ a Spectrum Technologies Watchdog 2700 weather station was used to record wind speed, direction, rainfall, temperature and relative humidity at one-hour intervals.
This allows end-users to identify what was used and identify the methods used more accurately.
Just as importantly, this acknowledges the contributions of others whose works were instrumental in your research.
This also helps ensure that you as a researcher can reconstruct what you have done since you will have good notes and documentation and be able to identify if something changes, _e.g._ a package version, what effect it had on your research.

The use of programming or scripting languages such as R, SAS or Python enables you as a scientist to keep very detailed records of what was computationally performed.
This is as opposed to using software such as spreadsheet programs like Excel, Google Sheets, Numbers, Calc or others which can be used for simple statistical analyses and visualization (but should not be) or other point-and-click software packages that do not enable you to keep an accurate record of the steps taken to import, format, visualise and analyse data.

@@

If steps are followed to make the data FAIR then it will be readable by humans and machines alike this will help support discoveries and support further research.
In turn, sharing data will lead to new citations for your work as others discover and use it.
